From Predicate Automata to Alternating Data Automata

### 2017-05-16 predicate-automata.tex
ok

% \documentclass{llncs}
\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}

@@

\usepackage{latexsym}
\usepackage{amsxtra} 
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{txfonts}
\usepackage{framed}
\usepackage{makecell}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{automata,positioning, calc}
\usepackage[inline,shortlabels]{enumitem}

@@

\title{The Impact of Alternation}

\author{Radu Iosif and Xiao Xu \\
CNRS, Verimag, Universit\'e de Grenoble Alpes \\
Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}

@@

\begin{abstract}
Alternating automata have been widely used to model and verify systems
that handle data from finite domains, such as communication protocols
or hardware. The main advantage of the alternating model of
computation is that complementation is possible in linear time, thus
allowing to concisely encode trace inclusion problems that occur often
in verification. In this paper we consider alternating automata over
infinite alphabets, whose transition rules are formulae in a combined
theory of booleans and some infinite data domain, that relate past and
current values of the data variables. The data theory is not fixed,
but rather it is a parameter of the class. We show that union,
intersection and complementation are possible in linear time in this
model and, though the emptiness problem is undecidable, we provide two
efficient semi-algorithms, inspired by two state-of-the-art
abstraction refinement model checking methods: lazy predicate
abstraction \cite{HJMS02} and the \impact~ semi-algorithm
\cite{mcmillan06}. We have implemented both methods and report the
results of an experimental comparison.
\end{abstract}

\section{Introduction}

The language inclusion problem is recognized as being central to
verification of hardware, communication protocols and software
systems. A property is a specification of the correct executions of a
system, given as a set $\mathcal{P}$ of executions, and the
verification problem asks if the set $\mathcal{S}$ of executions of
the system under consideration is contained within $\mathcal{P}$.

This problem is at the core of widespread verification techniques,
such as automata-theoretic model checking \cite{VARDI94}, where
systems are specified as finite-state automata and properties defined
using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
this and other related verification techniques is the intractability
of language inclusion (PSPACE-complete for finite-state automata on
finite words over finite alphabets).

Alternation \cite{ChandraKozenStockmeyer81} was introduced as a
generalization of nondeterminism, introducing universal, in addition
to existential transitions. For automata over finite alphabets, the
language inclusion problem can be encoded as the emptiness problem of
an alternating automaton of linear size. Moreover, efficient
exploration techniques based on antichains are shown to perform well
for alternating automata over finite alphabets \cite{DeWulf08}.

Using finite alphabets for the specification of properties and models
is however very restrictive, when dealing with real-life computer
systems, mostly because of the following reasons. On one hand,
programs handle data from very large domains, that can be assumed to
be infinite (64-bit integers, floating point numbers, strings of
characters, etc.) and their correctness must be specified in terms of
the data values. On the other hand, systems must respond to strict
deadlines, which requires temporal specifications as timed
languages \cite{AlurDill94}.

Although being convenient specification tools, automata over infinite
alphabets lack the decidability properties ensured by finite
alphabets. In general, when considering infinite data as part of the
input alphabet, language inclusion is undecidable and, even
complementation becomes impossible, for instance, for timed automata
\cite{AlurDill94} or finite-memory register automata
\cite{Kaminski94}. In some cases, one can recover theoretical
decidability, by restricting the number of variables (clocks) in timed
automata to one \cite{OuaknineWorrell04}, or forbidding relations
between current and past/future values, as with symbolic automata
\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.

In this paper, we present a new model of alternating automata over
infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
event from a finite set and $\nu$ is a valuation of a finite set
$\vec{x}$ of variables that range over an infinite domain. We assume
that, at all times, the successive values taken by the variables in
$\vec{x}$ are an observable part of the language, in other words,
there are no hidden variables in our model. The transition rules are
specified by a set of formulae, in a combined first-order theory of
boolean control states and data, that relate past with present values
of the variables. We do not fix the data theory a~priori, but rather
consider it to be a parameter of the class.

A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
rewritings of the initial formula by substituting boolean states with
time-stamped transition rules. The word is accepted if the final
formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
false and all final states by true.

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{HJMS02} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

%% Finally, we prove termination in the empty case, when the transition
%% formulae\begin{inparaenum}[(i)]
%% \item contain only constants of the data sort and boolean functions of
%%   any arity, or they
%% %
%% \item constrain only the current but not the past values on the input,
%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
%% \end{inparaenum}

We have implemented the boolean operations and emptiness checking
semi-algorithms and carried out experiments with examples taken from
array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

\paragraph{Related Work}
Data languages and automata have been defined previously, in a
classical nondeterministic setting. For instance, Kaminski and Francez
\cite{Kaminski94} consider languages, over an infinite alphabet of
data, recognized by automata with a finite number of registers, that
store the input data and compare it using equality. Just as the timed
languages recognized by timed automata \cite{AlurDill94}, these
languages, called quasi-regular, are not closed under complement, but
their emptiness is decidable. The impossibility of complementation
here is caused by the use of hidden variables, which we do not
allow. Emptiness is however undecidable in our case, mainly because
counting (incrementing and comparing to a constant) data values is
allowed, in many data theories.

%% Other approaches to defining classes of data automata
%% \cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
%% mainly because these automata are controlled by equivalence relations
%% of finite index, whereas in our case, the transitions are defined by
%% unrestricted formulae in the first-order theory of the data domain. As
%% in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
%% for these classes.

Another related model is that of predicate automata \cite{Farzan15},
which recognize languages over integer data by labeling the words with
conjunctions of uninterpreted predicates. We intend to explore further
the connection with our model of alternating data automata, in order
to apply our method to the verification of parallel programs.

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the decision procedure.

The emptiness check for alternating data automata adapts similar
semi-algorithms for nondeterministic infinite-state programs to the
alternating model of computation. In particular, we considered the
state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
outperform lazy predicate abstraction \cite{HJMS02} in the
nondeterministic case, and generalized it to cope with
alternation. More recent approaches for interpolant-based abstraction
refinement target Horn systems \cite{McMillan14,Hoder12}, used to
encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
cannot be encoded by a finite number of local (equality)
constraints. We believe that the lazy annotation techniques for Horn
clauses are suited for branching computations, which we intend to
consider in a future tree automata setting.

\section{Preliminaries}

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
  generalization to more than two sorts is without difficulty, but
  would unnecessarily clutter the technical presentation.} and each
function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
$\top$ and $\bot$ of sort $\Bool$.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. A
\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{formula} is an
existentially quantified combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
$t$. For indexed sets $\vec{t}=\set{t_1,\ldots,t_n}$ and
$\vec{x}=\set{x_1,\ldots,x_n}$, we write $\phi[\vec{t}/\vec{x}]$ for
the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
symbols occuring in $\phi$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.

A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A \emph{decision
procedure} for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

\section{Alternating Data Automata}

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
set of \emph{data symbols}. A \emph{data word} (word in the sequel) is
a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
words over $\vec{x}$.

This definition generalizes the classical notion of words from a
finite alphabet to the possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
$\Bool$ and $\Data$, respectively, we denote by
$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
under an even number of negations.

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
  configuration},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function},
\end{compactitem}
where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
each formula $\Delta(q,a)$ describing a transition rule, the variables
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe that the initial values of
the variables are left unconstrained, as the initial configuration
does not contain free data variables. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

\begin{figure}[htb]
\vspace*{-\baselineskip}
\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
\vspace*{-\baselineskip}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any
$k\geq0$, be a set of time-stamped variables. For an input event $a
\in \Sigma$ and a formula $\phi$, we write $\Delta(\phi,a)$
(respectively $\Delta^k(\phi,a)$) for the formula obtained from $\phi$
by simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
$k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
for some valuation $\nu$ that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.

\paragraph{\em Example} 
The following sequence is a non-accepting run of the ADA from Figure
\ref{fig:ex} on the word
$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$, where $\Data^\I =
\zed$ and the function symbols have standard arithmetic
interpretation:
\[\begin{array}{l}
q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
\underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
\end{array}\]

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}

It is well known that other problems, such as \emph{universality}
(given automaton $\A$ with variables $\vec{x}$, does $L(\A) =
\Sigma[\vec{x}]^*$?) and \emph{inclusion} (given automata $\A_1$ and
$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
L(\A_2)$?) can be reduced to the above problems. Observe furthermore
that we do not consider cases in which the sets of variables in the
two automata differ. An interesting problem in this case would be: given
automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
as future work.

\subsection{Boolean Closure}

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
\in \Form^+(Q,\vec{x})$.

In the following let $\A_i =
\tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, where
w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define:
\[\begin{array}{rcl}
\A_\cup & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup F_2,\Delta_1\cup\Delta_2} \\ 
\A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
\end{array}\] 
where $\overline{\Delta_1}(q,a) \equiv \overline{\Delta_1(q,a)}$, for
all $q \in Q_1$ and $a \in \Sigma$. The following lemma shows the
correctness of the above definitions:

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This matches the linear-time bounds for intersection and
complementation of alternating automata over finite alphabets
\cite{ChandraKozenStockmeyer81}.

\section{Antichains and Interpolants for Emptiness}

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) using only atoms $x \geq k$
and $x = \overline{x} + k$, for $k \in \zed$, with the classical
interpretation of the function symbols on integers. Since reachability
of a control state is undecidable for AVASS \cite{LINCOLN92}, the
emptiness problem is undecidable for ADA.

Consequently, we give up on the guarantee for termination and build
semi-algorithms that meet the requirements below: \begin{compactenum}[(i)]
\item given an automaton $\A$, if $L(\A) \neq\emptyset$, the procedure
  will terminate and return a word $w \in L(\A)$, and
%
\item if the procedure terminates without returning such a word, then
  $L(\A) = \emptyset$.
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading the event $a$.

We generalize the post-image function to finite sequences of input
events, as follows:
\[\begin{array}{l}
\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]
Then the emptiness problem for $\A$ becomes: does there exist $u \in
\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
Observe that, since we ask a satisfiability query, the final states of
$\A$ need not be constrained\footnote{ Since each state occurs
  positively in $\Accept{\A}(u)$, this formula has a model iff it has
  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
for $u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, even with this optimization, the enumeration of
sequences from $\Sigma^*$ diverges in many real cases, because
infinite antichains exist in many interpretations, e.g.\ $q \wedge x
\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
  $\A$ has a safety invariant.
\end{lemma}

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
defines a safety invariant for $\A$. Although this condition can be
effectively checked using a decision procedure for the theory
$\theory(\sig,\I)$, there is no guarantee that this check will ever
succeed.

The solution we adopt in the sequel is abstraction to ensure the
termination of invariant computations. However, it is worth pointing
out from the start that abstraction alone will only allow us to build
invariants that are not necessarily safety invariants. To meet the
latter condition, we resort to counterexample guided abstraction
refinement (CEGAR).

Formally, we fix a set of formulae $\preds \subseteq
\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
formulae as \emph{predicates}. Given a formula $\phi$, we denote by
$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
$\preds$. The abstract versions of the post-image and acceptance
condition are defined as follows:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
\end{array}\]

\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant $\abs{\Inv}$ for $\A$. 
\end{lemma}

>>>>

\documentclass{llncs}

@@

\usepackage{latexsym}
\usepackage{amsxtra} 
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pslatex}
% \usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{txfonts}
\usepackage{framed}
\usepackage{makecell}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{automata,positioning, calc}
\usepackage[inline,shortlabels]{enumitem}

@@

\title{From Predicate Automata to Alternating Data Automata}

\author{Radu Iosif}
\institute{CNRS, Verimag \\
Radu.Iosif@univ-grenoble-alpes.fr}

@@

\section{Definitions}

Let $Q$ be a finite set of predicates, where for each $q \in Q$, we
denote by $\#(q)\geq0$ its arity. We denote by $Q_0$ the subset of $Q$
consisting of predicates of arity zero. By $\Sigma$ we denote a
finite set of symbols, called alphabet.

We consider an infinite countable set $\vars$ of first-order
variables, ranging over $\nat$, and we associate each predicate $q \in
Q$ a tuple of arguments $\vec{x}_q = \tuple{x_{q,1}, \ldots,
  x_{q,\#(q)}}$ such that, for any two $p,q \in Q$, $\vec{x}_p$ and
$\vec{x}_q$ have no elements in common, when $p \neq q$.

\section{Translation of PA into ADA}
